Classical Planning with Simulators: \\ Results on the Atari Video Games

### 2015-02-09 ijcai15-paper/paper.tex
Miguel's Iteration #1
* Added Nir's sentence, with a few changes
* Fixed the bleeding into the References page following the above by
  trimming the space between the bullet points describing IW (via
  the package enumitem).
* Hunted a couple cases we were using the English British/Australian
  spelling for words ending on 'ze' (optmize vs. optimise, analyze vs.
  analyze, etc.).

\usepackage{todonotes}
\newcommand{\nir}{\todo[inline,color=orange!10]}
\newcommand{\hg}{\todo[color=blue!10]}
\newcommand{\mrj}{\todo[inline,color=green!10]}

@@

\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\frenchspacing
% \setlength{\pdfpagewidth}{8.5in}
% \setlength{\pdfpageheight}{11in}

@@

\author{\textbf{ID:} \ldots  \ \ ; \ \ \textbf{Keywords:} classical planning, search, video games}

@@

The action selection problem in the Atari games can be seen as a \emph{reinforcement learning} problem
\cite{sutton:book} over a deterministic MDP where the state transitions and rewards are not known, or alternatively,
as a  \emph{net-benefit planning problem} \cite{coles2012survey,keyder:jair09} with unknown state transitions  and  rewards. 
\Omit{Namely}
\nir{While the \emph{reinforcement learning} problem seeks to synthesize a controller as a policy/function that maximizes 
the reward of any trajectory}, we seek  an action sequence $a_0, a_1, \ldots, a_m$ from the  current  state $s_0$ that generates
a state sequence $s_0, s_1, \ldots, s_{m+1}$ with  maximum  total reward $\sum_{i=0}^{m} r(a_i,s_i)$,
where $m$ is a sufficiently large planning horizon, and the rewards $r(a_i,s_i)$ and the  state-transitions $s_{i+1}=f(a_i,s_i)$
are known only after  action $a_i$ is applied in the state $s_i$.

@@

The limitation of the   basic blind search  methods is that they are not effective  over large  spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line.
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planners: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is
called \IW\ for Iterated Width search \cite{nir:ecai2012}. \IW\ consists of a sequence of calls  $\IW(1)$, $\IW(2)$, .., $\IW(k)$, where
$\IW(i)$ is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set)
of at most $i$ atoms. Namely,   $\IW(1)$ is a breadth-first search that keeps  a state only if  the state is   the first  one
in the search to make some atom  true; $\IW(2)$ keeps a state  only if the state is  the first one to  make  a pair of atoms true, and
so on.  Like plain breadth-first  and iterative deepening searches,  \IW\  is
complete, while searching the state space in a way that makes use of the
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.
The  empirical results over \numgames\ Atari games show that  $\IW(1)$ performs at the level of UCT, 
the state-of-the-art planning method in this domain, and suggest the potential of width-based methods for planning with
simulators when factored, compact action models are not available.

@@

\begin{itemize}
\item The \emph{novelty of a newly generate state $s$  in a  search algorithm}
is   $1$ if  $s$ is the first state generated in the search that makes true some  atom $X=x$,
else it is $2$ if  $s$ is the first state that makes a \emph{pair} of atoms $X=x$ and $Y=y$ true,
and so on.

\item  $\IW(i)$ is a breadth-first search that prunes newly generated states when their novelty measure
is greater than $i$.

@@

For classical planning, the termination condition is the achievement of the goal. In the \emph{on-line setting,}
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by \IW \ is then the  path  that has a  maximum accumulated reward. The accumulated reward $R(s)$  
of a state $s$ reached in an iteration of \IW \ is determined by the  unique parent state $s'$
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by \IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected  in the on-line setting  is  the first action along  such a path. This action is then executed
and the process repeats from the resulting state.

@@

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain
relevant  in  optimisation problems as well. Indeed, if a  good path is made of
states $s_i$ each of which has a low width, \IW\ can be made
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in \IW\ to enforce this property.
% which is actually  not yet implemented in the code that is tested below.

@@

We tested \IW(1) and \BFS\ over \numgames\  of the $55$ different games 
considered in \cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
We left out  \textsc{Skiing} as the reported figures apparently use a  different reward structure.}
%% check
% \footnote{
% Entries marked as  ``--'' in the table are   still running at the time of submission. 
% Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
% Eventually all of  the  $55$ games will  be included in the evaluation.}
The two  algorithms were used to play the  games in the \emph{on-line planning
setting} supported by ALE where we will compare them with the planning algorithms 
considered by BNVB; namely, breadth-first search and \UCT. ALE supports also a \emph{learning setting} 
where the goal is to learn controllers that map  states into actions without
doing any lookahead. Algorithms across  the two settings are thus not directly comparable
as they compute different things. Learning controllers appears as a more challenging problem
and it is thus not surprising that planning algorithms like UCT tend to achieve a higher score than 
learning algorithms. In addition, the learning algorithms reported by BNVB tend to use 
the state of the screen pixels, while the planning algorithms,  use the state
of the RAM memory. It is not clear however whether one input representation is
more challenging than the other; for the learning algorithms, BNVB mention
that the results tend to be better for the screen inputs.

@@

\Omit{
The last difference inherited from BNVB between all algorithms
and \UCT\, is that \UCT\ selects an action randomly in the rollout every
single frame, while the others select an action every 5 frames, and
the remain ones are $no-op$, as is done in the simulation
environment. Therefore, a node in all algorithms represent $5$ frames,
but in UCT represents $1$ frame. As a result, a UCT rollout of depth
$300$ frames involve the selection of 300 actions, decreasing the
budget of simulated frames by $300$, while the other algorithms need a
branch of $1500$ frames to select $300$ actions, decreasing the budget
by $1500$ frames instead.
}

@@

 \caption{\small Performance that results from various  lookahead algorithms in \numgames\  Atari games. The algorithms, \BRFS, $\IW(1)$, \BFS, and \UCT, are evaluated
 over $10$ runs (episodes) for each game. 
The maximum episode duration is  $18,000$  frames and every algorithm is limited to a lookahead budget of 150,000 simulated frames.
Figures for \BRFS\ and \UCT\ taken from Bellemare et al. Average  CPU times per action in seconds, rounded to nearest integer, shown for  $\IW(1)$ and \BFS. 
Numbers  in bold show best performer in terms of average score, while numbers shaded in light grey show scores  that are better than \UCT's. 
Bottom part of the table shows pairwise comparisons among the algorithms.}
\label{performance}
\end{table*}

@@

In the iterated width algorithm, the key notion is the \emph{novelty measure} of a state in the underlying
breadth-first search. These novelty measures make use of the factored representation of the states for
providing a structure to the search: states that have width 1 are explored first in linear time,
then states that have  width 2  are explored in quadratic time, and so on. In classical planning problems
with atomic goals, this way of organizing the search  pays off both theoretically and practically.

>>>>

\usepackage{todonotes}
\newcommand{\nir}{\todo[color=orange!10]}
\newcommand{\hg}{\todo[color=blue!10]}
\newcommand{\mrj}{\todo[inline,color=green!10]}

@@

\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\frenchspacing
% \setlength{\pdfpagewidth}{8.5in}
% \setlength{\pdfpageheight}{11in}
\usepackage{enumitem}
%\setlist{nosep} % or \setlist{noitemsep} to leave space around whole list

@@

\author{\textbf{ID:} 320  \ \ ; \ \ \textbf{Keywords:} classical planning, search, video games}

@@

The action selection problem in the Atari games can be seen as a \emph{reinforcement learning} problem
\cite{sutton:book} over a deterministic MDP where the state transitions and rewards are not known, or alternatively,
as a  \emph{net-benefit planning problem} \cite{coles2012survey,keyder:jair09} with unknown state transitions  and  rewards. 
%\Omit{Namely} \Omit{While the \emph{reinforcement learning} problem seeks to synthesize a controller as a policy/function that maximizes 
%the reward of any trajectory\nir{NIR remark}}
While in the former a policy mapping every state $s$ to an action $a$ that maximizes expected reward is sought, in the latter one 
seeks an action sequence $a_0, a_1, \ldots, a_m$ from the  \emph{current}  state $s_0$ that generates
a state sequence $s_0, s_1, \ldots, s_{m+1}$ with  maximum  total reward $\sum_{i=0}^{m} r(a_i,s_i)$,
where $m$ is a sufficiently large planning horizon, and the rewards $r(a_i,s_i)$ and the  state-transitions $s_{i+1}=f(a_i,s_i)$
are known only after  action $a_i$ is applied in the state $s_i$.

@@

The limitation of the   basic blind search  methods is that they are not effective  over large state spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line.
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planners: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is
called \IW\ for Iterated Width search \cite{nir:ecai2012}. \IW\ consists of a sequence of calls  $\IW(1)$, $\IW(2)$, .., $\IW(k)$, where
$\IW(i)$ is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set)
of at most $i$ atoms. Namely,   $\IW(1)$ is a breadth-first search that keeps  a state only if  the state is   the first  one
in the search to make some atom  true; $\IW(2)$ keeps a state  only if the state is  the first one to  make  a pair of atoms true, and
so on.  Like plain breadth-first  and iterative deepening searches,  \IW\  is
complete, while searching the state space in a way that makes use of the
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.
The  empirical results over \numgames\ Atari games show that  $\IW(1)$ performs at the level of UCT, 
the state-of-the-art planning method in this domain, and suggest the potential of width-based methods for planning with
simulators when factored, compact action models are not available.

@@

\begin{itemize}[noitemsep]
\item The \emph{novelty of a newly generate state $s$  in a  search algorithm}
is   $1$ if  $s$ is the first state generated in the search that makes true some  atom $X=x$,
else it is $2$ if  $s$ is the first state that makes a \emph{pair} of atoms $X=x$ and $Y=y$ true,
and so on.

\item  $\IW(i)$ is a breadth-first search that prunes newly generated states when their novelty \Omit{measure}
is greater than $i$.

@@

\medskip
For classical planning, the termination condition is the achievement of the goal. In the \emph{on-line setting,}
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by \IW \ is then the  path  that has a  maximum accumulated reward. The accumulated reward $R(s)$  
of a state $s$ reached in an iteration of \IW \ is determined by the  unique parent state $s'$
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by \IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected  in the on-line setting  is  the first action along  such a path. This action is then executed
and the process repeats from the resulting state.

@@

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain
relevant  in  optimization problems as well. Indeed, if a  good path is made of
states $s_i$ each of which has a low width, \IW\ can be made
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in \IW\ to enforce this property.
% which is actually  not yet implemented in the code that is tested below.

@@

We tested \IW(1) and \BFS\ over \numgames\  of the $55$ different games 
considered in \cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
We left out  \textsc{Skiing} as the reported figures apparently use a  different reward structure.}
%% check
% \footnote{
% Entries marked as  ``--'' in the table are   still running at the time of submission. 
% Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
% Eventually all of  the  $55$ games will  be included in the evaluation.}
The two  algorithms were used to play the  games in the \emph{on-line planning
setting} supported by ALE where we will compare them with the planning algorithms 
considered by BNVB; namely, breadth-first search and \UCT. ALE supports also a \emph{learning setting} 
where the goal is to learn controllers that map  states into actions without
doing any lookahead. Algorithms across  the two settings are thus not directly comparable
as they compute different things. Learning controllers appears as a more challenging problem
and it is thus not surprising that planning algorithms like \UCT\ tend to achieve a higher score than 
learning algorithms. In addition, the learning algorithms reported by BNVB tend to use 
the state of the screen pixels, while the planning algorithms,  use the state
of the RAM memory. It is not clear however whether one input representation is
more challenging than the other; for the learning algorithms, BNVB mention
that the results tend to be better for the screen inputs.

@@

\Omit{
The last difference inherited from BNVB between all algorithms
and \UCT\, is that \UCT\ selects an action randomly in the rollout every
single frame, while the others select an action every 5 frames, and
the remain ones are $no-op$, as is done in the simulation
environment. Therefore, a node in all algorithms represent $5$ frames,
but in UCT represents $1$ frame. As a result, a \UCT\ rollout of depth
$300$ frames involve the selection of 300 actions, decreasing the
budget of simulated frames by $300$, while the other algorithms need a
branch of $1500$ frames to select $300$ actions, decreasing the budget
by $1500$ frames instead.
}

@@

 \caption{\small Performance that results from various  lookahead algorithms in \numgames\  Atari 2600 games. The algorithms, \BRFS, $\IW(1)$, \BFS, and \UCT, are evaluated
 over $10$ runs (episodes) for each game. 
The maximum episode duration is  $18,000$  frames and every algorithm is limited to a lookahead budget of 150,000 simulated frames.
Figures for \BRFS\ and \UCT\ taken from Bellemare et al. Average  CPU times per action in seconds, rounded to nearest integer, shown for  $\IW(1)$ and \BFS. 
Numbers  in bold show best performer in terms of average score, while numbers shaded in light grey show scores  that are better than \UCT's. 
Bottom part of the table shows pairwise comparisons among the algorithms.}
\label{performance}
\end{table*}

@@

In the iterated width algorithm, the key notion is the \emph{novelty measure} of a state in the underlying
breadth-first search. These novelty measures make use of the factored representation of the states for
providing a structure to the search: states that have width $1$ are explored first in linear time,
then states that have  width $2$ are explored in quadratic time, and so on. In classical planning problems
with atomic goals, this way of organizing the search  pays off both theoretically and practically.
